---
title: "STA2201 Week 1 Lab"
author: "Yeonjoon Choi"
date: "2023-01-11"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab Exercises

1. Plot the ratio of male to female mortality rates over time for ages 10,20,30 and 40 (different color for each age) and change the theme 
2. Find the age that has the highest female mortality rate each year 
3. Use the `summarize(across())` syntax to calculate the standard deviation of mortality rates by age for the Male, Female and Total populations. 
4. The Canadian HMD also provides population sizes over time (https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt). Use these to calculate the population weighted average mortality rate separately for males and females, for every year. Make a nice line plot showing the result (with meaningful labels/titles) and briefly comment on what you see (1 sentence). Hint: `left_join` will probably be useful here. 


Part 1.

```{r}
require(tidyverse)
dm = read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", 
                skip = 2, col_types = "dcddd")


#We get the mortality ratio for each age
data =  dm |>
    filter(Age == 10|Age == 20| Age == 30| Age ==40)|>
    select(Year:Male)|>
    mutate(ratio = Male/Female)

data

#We plot the data
data |>
  ggplot(aes(Year, ratio))+
  geom_point(aes(colour = Age))+
  theme_bw()+
  labs(title = "Ratio of Male-to-Female Mortality, from 1921 to 2019")+
  ylab ("Mortality Ratio")


```

Part 2.

```{r}
#We group by Year first
grouped_year = dm |>
  group_by(Year)

#and find the age that has
#the max mortality each year
mortality_age = summarize(grouped_year, 
          Max_Age=Age[which.max(Female)])


#We will see the first 10 years and the last 10 years 
head(mortality_age, n =10)

tail(mortality_age, n =10)


```


Part 3

```{r}

#We group by age
grouped_age = dm |>
  group_by(Age)

#and calculate the standard deviation
summarize(grouped_age, across(Female:Total, sd))
          

```



Part 4

```{r}
pop_data = read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt",
                      skip = 2)
death_data = read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", 
                        skip = 2, col_types = "dcddd")

death_data = death_data|>
  select(-Total)|>
  pivot_longer(Female:Male, names_to = "sex", values_to = "mortality")

#death data has no 2020 data
#so we remove any data after 2020
#from the data set
pop_data = pop_data|>
  select(-Total)|>
  filter(Year < 2020)|>
  pivot_longer(Female:Male, names_to = "sex", values_to = "Population")


death_data$population = pop_data$Population

#Calculate the weight for the weighted average
death_data = death_data |>
  mutate(weight = mortality*population)


#Calculate weighted average

to_plot = death_data |>
  group_by(Year, sex) |>
  summarize(weighted_mean = sum(weight, na.rm = TRUE)/sum(population, na.rm = TRUE))


#Plotting 

to_plot |>
ggplot(aes(Year, weighted_mean, colour = sex))+
 geom_point()+
  geom_line()+
  labs(title = "Mortality Rate for Female and Male, 1921-2019")+
  ylab("Mortality Rate")+
  theme_bw()

  
```

We see that male mortality rate is higher than female mortality rate, but we also see that the difference shrunk as time went on. 